Data Analysis

### 2025-08-31 code/07_data_analysis.Rmd
Code checked & formatted; html rendered

# Loading Data for analysis

```{r load_packages-data, warning=FALSE, message=FALSE}

# Setting up workspace

# For plotting trees, and although not strictly necessary, we may need the `ggtree` R package, which you should be able to install by running:
 if (!require("BiocManager", quietly = TRUE))
 install.packages("BiocManager")
 BiocManager::install("ggtree")

if (!require("orchaRd", quietly = TRUE))
devtools::install_github("daniel1noble/orchaRd", ref = "main", force = TRUE)

pacman::p_load(here,tidyverse,metafor,ggpubr,orchaRd,ggpubfigs)

#Packages needed

rm(list=ls()) ## cleaning up

set.seed(430) #set seed for reproducibility (for random number generation)

# Functions needed

# function to estimate typical sampling error variance obtained from 
# https://github.com/Yefeng0920/heterogeneity_ecoevo/tree/main/function
sigma2_v <- function(mod){
  sigma2_v <- sum(1 / mod$vi) * (mod$k - 1) /
    (sum(1 / mod$vi)^2 - sum((1 / mod$vi)^2))
  return(sigma2_v)
}

@@

dataset_analysis<-read_csv(here::here("data/04_data_analysis/dataset_analysis.csv"))

dataset_lnRR<-read_csv(here::here("data/04_data_analysis/dataset_lnRR.csv"))

dataset_SMDH<-read_csv(here::here("data/04_data_analysis/dataset_SMDH.csv"))

### Creating Phylogeny for the plot

```{r}
# install.packages("pacman")
# load packages

pacman::p_load(rotl,
               ape,
               ggtree)

# We have 7 bird species in the dataset
bird_species <- unique(dataset_analysis$bird_species)

# extracting taxonomic information
taxa <- rotl::tnrs_match_names(names = bird_species)
taxa

# check approximate matches
taxa[taxa$approximate_match==TRUE & !(is.na(taxa$approximate_match)),]

# phylogenetic information for our species from the Open Tree of Life

tree <- rotl::tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name")
ape::is.binary(tree)

# here are the species included in the tree
sort(tree$tip.label)

# removing the underscore "_" from the tree tip.label
tree$tip.label <- gsub("_"," ", tree$tip.label)
sort(tree$tip.label)

# compute branch lengths of tree
phylo_branch <- compute.brlen(tree, method = "Grafen", power = 1)

# check tree is ultrametric
is.ultrametric(phylo_branch) # TRUE

# matrix to be included in the models
bird_phylo <- vcv(phylo_branch, cor = T)
bird_phylo

plot(tree) #check: https://www.rdocumentation.org/packages/ape/versions/5.3/topics/plot.phylo
```

### Visualizing random effects structure

```{r}
# install.packages("devtools")
devtools::install_github("davidsjoberg/ggsankey")
library(ggsankey)

sankey_data <- dataset_analysis %>%
  select(Observation_ID, repeated_trait_ID_coded, paper_ID, experiment_ID_coded, group_ID_coded)%>%
  mutate(across(c(Observation_ID, repeated_trait_ID_coded, paper_ID, experiment_ID_coded, group_ID_coded), as.factor)) %>%
  make_long(Observation_ID, repeated_trait_ID_coded, paper_ID, experiment_ID_coded, group_ID_coded)

sankey_data <- dataset_analysis %>%
  select(paper_ID, bird_species, population_ID)%>%
  mutate(across(c(paper_ID, bird_species, population_ID), as.factor)) %>%
  make_long(paper_ID, bird_species, population_ID)

ggplot(sankey_data, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               fill = factor(node))) +
  geom_sankey() +
  scale_fill_discrete(drop=FALSE) +
  theme(legend.position = "none")

ggplot(sankey_data, aes(x = x, next_x = next_x, node = node, next_node = next_node, fill = factor(node))) +
  geom_alluvial(flow.alpha = .6) +
  scale_fill_viridis_d(drop = FALSE) +
  theme_alluvial(base_size = 18) +
  labs(x = NULL) +
  theme(legend.position = "none",
        plot.title = element_text(hjust = .5))

# Preparing datasets

```{r}
# Let's just quickly ensure all the important variables are 
# in the correct data type

dataset_analysis<- dataset_analysis%>%mutate(across(c(paper_ID, Observation_ID, repeated_trait_ID_coded, paper_ID, experiment_ID_coded, group_ID_coded, comparision_type,Hypothesis,trait_type,parasite_type,time_of_gnm_addition, blinding, random_assignment, missing_data, population_ID,bird_species), as.factor))

dataset_lnRR<- dataset_lnRR%>%mutate(across(c(paper_ID, Observation_ID, repeated_trait_ID_coded, paper_ID, experiment_ID_coded, group_ID_coded, comparision_type,Hypothesis,trait_type,parasite_type,time_of_gnm_addition, blinding, random_assignment, missing_data, population_ID, bird_species), as.factor))

dataset_SMDH<- dataset_SMDH%>%mutate(across(c(paper_ID, Observation_ID, repeated_trait_ID_coded, paper_ID, experiment_ID_coded, group_ID_coded, comparision_type,Hypothesis,trait_type,parasite_type,time_of_gnm_addition, blinding, random_assignment, missing_data, population_ID,bird_species), as.factor))

```{r seperating datasets for analysis}

## For lnRR (or simply load dataset as provided)
# dataset_lnRR<-dataset_analysis%>%
# filter(lnRR_sign != "NA" | lnRR_variance != "NA")

# selecting dataset which are not flagged
dataset_lnRR<-dataset_lnRR%>%
  filter(proxy_decision == "include")

# selecting only published dataset
dataset_lnRR_published<-dataset_lnRR%>%
  filter(!is.na(year_publication))

# selecting dataset which are flagged
lnRR_con_flagged<-dataset_analysis%>%
   filter(lnRR_sign != "NA" | lnRR_variance != "NA")%>%
  filter(proxy_decision != "SMDH only" | proxy_decision != "contingency table" )

# selecting dataset without those using 0 as ES because authors reported no effect
lnRR_sin_missingES <- dataset_lnRR %>% 
   filter(!proxy_comment %in% c("use 0 as ES"))

lnRR_flagged_no_missingES<- lnRR_con_flagged %>% 
   filter(!proxy_comment %in% c("use 0 as ES"))

# selecting only dataset that passes geary's test
lnRR_geary_passed<-dataset_lnRR%>%filter(geary_test=="pass")

## For SMDH
# selecting dataset which are not flagged
dataset_SMDH<-dataset_analysis%>%
   filter(proxy_decision == "include"| proxy_decision == "SMDH only"|proxy_decision =="contingency table") %>% filter(SMDH_sign != "NA")

# selecting only published dataset
dataset_SMDH_published<-dataset_SMDH%>%
  filter(!is.na(year_publication))

# selecting dataset which are flagged
SMDH_con_flagged<-dataset_analysis%>%
   filter(SMDH_sign != "NA")

# selecting dataset without those using 0 as ES because authors reported no effect

@@

SMDH_flagged_no_missingES<- SMDH_con_flagged %>% 
   filter(!proxy_comment %in% c("use 0 as ES"))

# selecting dataset which are not flagged but without the SMDH calculated from inferential statistics

SMDH_no_inferential<-dataset_analysis%>%
  filter(proxy_decision == "include")%>% filter(SMDH_sign != "NA")
```

# Overall effect : Intercept-only meta-analytic model

@@

## using lnRR

```{r intercept-only-model-lnRR}
VCV_lnRR<- vcalc(vi = lnRR_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_lnRR,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )

intercept_lnRR<- rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR, # specify lnRR's sampling variance-covariance matrix;
                    mods = ~ 1,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                    control = list(optimizer="optim"),
                    test = "t",
                    data = dataset_lnRR)
# saving the model
saveRDS(intercept_lnRR, file = here::here("model/intercept_lnRR.rds"))

# saving the dataset for this model
write.csv(dataset_lnRR,here::here("data/04_data_analysis/intercept_lnRR.csv"))
```

@@

```{r function-interpret-heterogeneity}
# ## Function is from the Pluralistic approach to heterogeneity from Yefeng's Pre-print repository from GitHub
# 
# #-----------------------heterogeneity interpretation-----------------------#
# het_interpret <- function(observed_value, het_type, es_type, data) {
#   
#   if (!het_type %in% c("I2", "CVH", "M", "sigma2", "V_bar")) {
#     stop("Invalid heterogeneity type. Choose from 'I2', 'CVH', 'M', 'sigma2', or 'V_bar'.")
#   }
#   
#   if (!es_type %in% unique(data$es.type)) {
#     stop("Invalid effect size type. Check the levels of 'es.type'.")
#   }
#   
#   filtered_data <- data %>% filter(es.type == es_type)
#   
#   if (nrow(filtered_data) == 0) {
#     stop("No data available for the selected effect size type.")
#   }
#   
#   het_values <- filtered_data[[het_type]]
#   
#   percentiles <- quantile(het_values, probs = seq(0, 1, by = 0.05), na.rm = TRUE)
#   
#   lower_bound <- max(percentiles[percentiles <= observed_value], na.rm = TRUE)
#   upper_bound <- min(percentiles[percentiles >= observed_value], na.rm = TRUE)
#   lower_percentile <- names(percentiles)[percentiles == lower_bound]
#   upper_percentile <- names(percentiles)[percentiles == upper_bound]
#   
#   percentile_range <- paste0(lower_percentile, "-", upper_percentile, "th percentile")
#   
#   # return the results
#   return(tibble::tibble(
#     observed_value = observed_value,
#     het_type = het_type,
#     es_type = es_type,
#     percentile_range = percentile_range
#   ))
# }
# x<-het_interpret(observed_value= I2_main_model_vector[1], 
#                  het_type = "I2", 
#                  es_type = "lnRR_sign", 
#                  data= intercept_lnRR)

@@

```{r}

# Then also:
# # Typical sampling variance (which captures the statistical noise of the data, 
# # but is rarely reported in the current meta-analytic practice):
sigma2_v(intercept_lnRR)

@@

# # # Total unstandardized raw variance (i.e. total heterogeneity)
round(sum(intercept_lnRR$sigma2), 3)

@@

```{r intercept-only-model-SMDH}

VCV_SMDH <- vcalc(vi = SMDH_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_SMDH,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )

intercept_SMDH<- rma.mv(yi = SMDH_sign, # specify SMDH as the effect size measure;
                    V = VCV_SMDH, # specify SMDH's sampling variance-covariance matrix;
                    mods = ~ 1,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                    control = list(optimizer="optim"),
                    test = "t",
                    data = dataset_SMDH)

# saving the model
saveRDS(intercept_SMDH, file = here::here("model/intercept_SMDH.rds"))

# saving the dataset for this model
write.csv(dataset_SMDH,here::here("data/04_data_analysis/intercept_SMDH.csv"))
```

## SMDH-heterogeneity

@@

# # # Total unstandardized raw variance (i.e. total heterogeneity)
round(sum(intercept_SMDH$sigma2), 3)

@@

```{r}
dataset_bivariate <- data.frame(paper_ID = c(dataset_lnRR$paper_ID, dataset_SMDH$paper_ID),
                      Observation_ID = c(dataset_lnRR$Observation_ID, dataset_SMDH$Observation_ID),
                      group_ID_coded = c(dataset_lnRR$group_ID_coded, dataset_SMDH$group_ID_coded),
                      population_ID = c(dataset_lnRR$population_ID, dataset_SMDH$population_ID),
                      bird_species = c(dataset_lnRR$bird_species, dataset_SMDH$bird_species),
                      ES = c(dataset_lnRR$lnRR_sign, dataset_SMDH$SMDH_sign),
                      V_ES = c(dataset_lnRR$lnRR_variance, dataset_SMDH$SMDH_variance),
                      ES_measure = c(rep("lnRR", nrow(dataset_lnRR)), rep("SMDH", nrow(dataset_SMDH)))
                      )

VCV_bivariate <- vcalc(vi = V_ES, # sampling variances of lnRR and SMD that are correlated within the same study;
             cluster = paper_ID, # study identity;
             type = ES_measure, # different types of effect size measures underlying the observed effect sizes;
             data = dataset_bivariate, # the long format data frame;
             rho = 0.2, # assuming that the effect sizes within the same study are correlated with rho = 0.5.
             nearpd = TRUE # specify whether impose the non positive definite VCV matrix (not invertible) to the nearest positive semi-definite matrix 
             )

Bivariate_model <- rma.mv(yi = ES, # specify the effect size estimate (the variable ES in our case);
                V = VCV_bivariate, # specify the imputed variance-covariance matrix;
                mods = ~ ES_measure - 1, # specify the variable "ES_measure" indicating the types of effect size measures;
                random = list(~ ES_measure | paper_ID, 
                              ~ ES_measure | Observation_ID,
                              ~ 1 | group_ID_coded,
                              ~ 1 | population_ID,
                              ~ 1 | bird_species
                              ), # add correlated random effects corresponding to the lnRR and SMD parameters in the same study;
                struct = "UN", # impose an unstructured variance-covariance of the study-specific random effects;
                test = "t", # t distribution is specified to test the overall effect against the null hypothesis and construct confidence intervals;
                method = "REML", # restricted likelihood maximum is assigned as the estimator for variance components as suggested;
                data = dataset_bivariate, # the long format dataset
                sparse = T
               )

# Saving the model
saveRDS(Bivariate_model, file = here::here("model/sensitivity_analysis/bivariate_model.rds"))
# saving the dataset for this model
write.csv(dataset_bivariate,here::here("data/05_data_sensitivity_analysis/dataset_bivariate.csv"))

@@

```{r WARNING = F}
pacman::p_load(ellipse)
ab.l <- matreg(y=2, x=1, R=Bivariate_model$G, cov=TRUE, means=coef(Bivariate_model), n=Bivariate_model$g.levels.comb.k) # fit regression model

xy <- ellipse(Bivariate_model$G, centre=coef(Bivariate_model), level=0.95) # get 95% coverage region
ellipse_df <- data.frame(x = xy[, 1], y = xy[, 2]) # convert it int dataframe so that we can use ggplot to make a figure

@@

There is a very high negative correlation between the two effect sizes. This probably is an artifact of the data and not a true correlation. If you look at the model summary, the $tau^2$ for lnRR, it is extremely small 0.0008, i.e., there is very little variance. This could also be stemming from the fact that the log values in the dataset could be causing this. For SMDH on the other hand, the $tau^2$ is 0.1681 that is much larger. This difference in the variance estimate itself could be causing this artifact of the model. I will only go ahead to present this model as a senstivity analysis because I am not completely convinced by it's robustness.

# Sensitivity Analysis

@@

```{r}

#----------------------------------------------------------------------------
# for lnRR as Effect Size
#----------------------------------------------------------------------------

VCV_lnRR_published <- vcalc(vi = lnRR_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_lnRR_published,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )

SA_lnRR_published<- rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR_published, # specify lnRR's sampling variance-covariance matrix;
                    mods = ~ 1,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                    control = list(optimizer="optim"),
                    test = "t",
                    data = dataset_lnRR_published)

# Saving the model
saveRDS(SA_lnRR_published, file = here::here("model/sensitivity_analysis/SA_lnRR_published.rds"))
# saving the dataset for this model
write.csv(dataset_lnRR_published,here::here("data/05_data_sensitivity_analysis/dataset_lnRR_published.csv"))

# What is the effect now?

summary(SA_lnRR_published)
predict(SA_lnRR_published)

>>>>

## Preface
### Setting up the workspace

* Install and load required packages

```{r setup, warning=FALSE, message=FALSE}
## Setting up workspace using pacman
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

# For plotting trees, and although not strictly necessary, 
# we may need the `ggtree` R package, which you should be able to install by running:
if (!requireNamespace("ggtree", quietly = TRUE)) {
  ## Install using BiocManager
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  BiocManager::install("ggtree")  
}

## installing orchaRd from GitHub
if (!requireNamespace("orchaRd", quietly = TRUE)) {
  ## installing orchaRd from GitHub
  pacman::p_install_gh("daniel1noble/orchaRd", ref = 'main')
}

## installing ggsankey from GitHub
if (!requireNamespace("ggsankey", quietly = TRUE)) {
  ## installing orchaRd from GitHub
  pacman::p_install_gh("davidsjoberg/ggsankey", ref = 'main')
}

## Install/load required packages
pacman::p_load(
  car,
  ellipse,
  here,
  ggpubfigs,
  ggpubr,
  ggsankey,
  metafor,
  orchaRd,
  tidyverse
)
```

### Custom functions

For the presented analyses, we need to define the function [sigma2_v](https://github.com/Yefeng0920/heterogeneity_ecoevo/blob/main/function/function.R) to estimate typical sampling error variance

```{r custom functions, warning=FALSE, message=FALSE}
sigma2_v <- function(mod){
  sigma2_v <- sum(1 / mod$vi) * (mod$k - 1) /
    (sum(1 / mod$vi)^2 - sum((1 / mod$vi)^2))
  return(sigma2_v)
}
```

### Loading Data for analysis

```{r load_packages-data, warning=FALSE, message=FALSE}
## set seed for reproducibility 
## (for random number generation)
set.seed(430)

@@

dataset_analysis <- read_csv(here::here("data/04_data_analysis/dataset_analysis.csv"))

dataset_lnRR <- read_csv(here::here("data/04_data_analysis/dataset_lnRR.csv"))

dataset_SMDH <- read_csv(here::here("data/04_data_analysis/dataset_SMDH.csv"))
```

## Creating Phylogeny for the plot

```{r phylogeny, eval=FALSE}
## NOT run
## -----------------------------------------------------------

## required packages
pacman::p_load(rotl, ape, ggtree)

## We have 7 bird species in the dataset
bird_species <- unique(dataset_analysis$bird_species)

## extracting taxonomic information
(taxa <- rotl::tnrs_match_names(names = bird_species))

## check approximate matches
taxa[taxa$approximate_match==TRUE & !(is.na(taxa$approximate_match)),]

## phylogenetic information for our species from the Open Tree of Life

tree <- rotl::tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name")
## check
ape::is.binary(tree)

## here are the species included in the tree
sort(tree$tip.label)

## removing the underscore "_" from the tree tip.label
tree$tip.label <- gsub("_"," ", tree$tip.label)
sort(tree$tip.label)

## compute branch lengths of tree
phylo_branch <- compute.brlen(tree, method = "Grafen", power = 1)

## check tree is ultrametric
is.ultrametric(phylo_branch)

## matrix to be included in the models
bird_phylo <- vcv(phylo_branch, cor = T)
bird_phylo

## plot(tree) #check: https://www.rdocumentation.org/packages/ape/versions/5.3/topics/plot.phylo
plot(tree)
```

## Visualizing random effects structure

```{r random-effects}
sankey_data <- dataset_analysis %>%
  ## select variables
  select(Observation_ID, repeated_trait_ID_coded, paper_ID, experiment_ID_coded, group_ID_coded) %>%
  mutate(across(c(Observation_ID, repeated_trait_ID_coded, paper_ID, experiment_ID_coded, group_ID_coded), as.factor)) %>%
  make_long(Observation_ID, repeated_trait_ID_coded, paper_ID, experiment_ID_coded, group_ID_coded)

sankey_data <- dataset_analysis %>%
  select(paper_ID, bird_species, population_ID) %>%
  mutate(across(c(paper_ID, bird_species, population_ID), as.factor)) %>%
  make_long(paper_ID, bird_species, population_ID)

## visualise sankey_data
ggplot(sankey_data,
       aes(x = x, next_x = next_x, node = node, next_node = next_node, fill = factor(node))) +
  geom_alluvial(flow.alpha = .6) +
  scale_fill_viridis_d(drop = FALSE) +
  theme_alluvial(base_size = 18) +
  labs(x = NULL) +
  theme(legend.position = "none",
        plot.title = element_text(hjust = .5))
```

## Preparing datasets

```{r}
# Let's just quickly ensure all the important variables are 
# in the correct data type
dataset_analysis<- dataset_analysis %>%
  mutate(across(c(paper_ID, Observation_ID, repeated_trait_ID_coded,
                  paper_ID, experiment_ID_coded, group_ID_coded,
                  comparision_type, Hypothesis, trait_type,
                  parasite_type, time_of_gnm_addition, blinding,
                  random_assignment, missing_data, population_ID,
                  bird_species),
                as.factor))

dataset_lnRR <- dataset_lnRR %>%
  mutate(across(c(paper_ID, Observation_ID, repeated_trait_ID_coded,
                  paper_ID, experiment_ID_coded, group_ID_coded,
                  comparision_type, Hypothesis, trait_type,
                  parasite_type, time_of_gnm_addition, blinding,
                  random_assignment, missing_data, population_ID, 
                  bird_species),
                as.factor))

dataset_SMDH <- dataset_SMDH %>% mutate(across(c(paper_ID, Observation_ID, repeated_trait_ID_coded,
                                                 paper_ID, experiment_ID_coded, group_ID_coded,
                                                 comparision_type, Hypothesis, trait_type,
                                                 parasite_type, time_of_gnm_addition, blinding,
                                                 random_assignment, missing_data, population_ID,
                                                 bird_species),
                                               as.factor))
```

```{r seperating datasets for analysis}
## For lnRR (or simply load dataset as provided)
dataset_lnRR <- dataset_analysis %>%
  filter(lnRR_sign != "NA" | lnRR_variance != "NA")

## selecting dataset which are not flagged
dataset_lnRR <- dataset_lnRR %>%
  filter(proxy_decision == "include")

## selecting only published dataset
dataset_lnRR_published <- dataset_lnRR %>%
  filter(!is.na(year_publication))

## selecting dataset which are flagged
lnRR_con_flagged <- dataset_analysis %>%
  filter(lnRR_sign != "NA" | lnRR_variance != "NA")%>%
  filter(proxy_decision != "SMDH only" | proxy_decision != "contingency table" )

## selecting dataset without those using 0 as ES because authors reported no effect
lnRR_sin_missingES <- dataset_lnRR %>% 
  filter(!proxy_comment %in% c("use 0 as ES"))

lnRR_flagged_no_missingES <- lnRR_con_flagged %>% 
  filter(!proxy_comment %in% c("use 0 as ES"))

## selecting only dataset that passes geary's test
lnRR_geary_passed <- dataset_lnRR %>%
  filter(geary_test == "pass")

## For SMDH
## --------------------------------------------------------------
## selecting dataset which are not flagged
dataset_SMDH <- dataset_analysis %>%
  filter(proxy_decision != "sensitivity-analysis" | is.na(proxy_decision))

## selecting only published datasets
dataset_SMDH_published <- dataset_SMDH %>%
  filter(!is.na(year_publication))

## selecting dataset which are flagged
SMDH_con_flagged <- dataset_analysis %>%
  filter(SMDH_sign != "NA")

## selecting dataset without those using 0 as ES 
## because authors reported no effect

@@

SMDH_flagged_no_missingES <- SMDH_con_flagged %>% 
  filter(!proxy_comment %in% c("use 0 as ES"))

## selecting dataset which are not flagged 
## but without the SMDH calculated from inferential statistics

SMDH_no_inferential <- dataset_analysis %>%
  filter(proxy_decision == "include") %>%
  filter(SMDH_sign != "NA")
```

## Overall effect : Intercept-only meta-analytic model

@@

### using lnRR

```{r intercept-only-model-lnRR}
VCV_lnRR <- vcalc(
  vi = lnRR_variance,
  cluster = paper_ID,
  obs = Observation_ID,
  data = dataset_lnRR,
  # assuming that the effect sizes within the same study 
  # are correlated with rho = 0.5
  rho = 0.5)

intercept_lnRR <- rma.mv(
  # specify lnRR as the effect size measure;
  yi = lnRR_sign,
  # specify lnRR's sampling variance-covariance matrix;
  V = VCV_lnRR, 
  mods = ~ 1,
  # set random effects structure
  random = list(~ 1 | paper_ID,
                ~ 1 | Observation_ID,
                ~ 1 | group_ID_coded,
                ~ 1 | population_ID,
                ~ 1 | bird_species),
  method = "REML",
  control = list(optimizer="optim"),
  test = "t",
  data = dataset_lnRR)
```

```{r, eval = FALSE}
## NOT RUN

## saving the model
saveRDS(intercept_lnRR, 
        file = here::here("model/intercept_lnRR.rds"))

## saving the dataset for this model
write.csv(dataset_lnRR,
          here::here("data/04_data_analysis/intercept_lnRR.csv"))
```

@@

```{r function-interpret-heterogeneity, eval=FALSE}
# Sun Aug 31 12:01:40 2025 ------------------------------
# Uncommented this code chunk but set eval = F

## Function is from the Pluralistic approach to heterogeneity 
## from Yefeng's Pre-print repository from GitHub

## -----------------------heterogeneity interpretation-----------------------#
het_interpret <- function(observed_value, het_type, es_type, data) {

  if (!het_type %in% c("I2", "CVH", "M", "sigma2", "V_bar")) {
    stop("Invalid heterogeneity type. Choose from 'I2', 'CVH', 'M', 'sigma2', or 'V_bar'.")
  }

  if (!es_type %in% unique(data$es.type)) {
    stop("Invalid effect size type. Check the levels of 'es.type'.")
  }

  filtered_data <- data %>% filter(es.type == es_type)

  if (nrow(filtered_data) == 0) {
    stop("No data available for the selected effect size type.")
  }

  het_values <- filtered_data[[het_type]]

  percentiles <- quantile(het_values, probs = seq(0, 1, by = 0.05), na.rm = TRUE)

  lower_bound <- max(percentiles[percentiles <= observed_value], na.rm = TRUE)
  upper_bound <- min(percentiles[percentiles >= observed_value], na.rm = TRUE)
  lower_percentile <- names(percentiles)[percentiles == lower_bound]
  upper_percentile <- names(percentiles)[percentiles == upper_bound]

  percentile_range <- paste0(lower_percentile, "-", upper_percentile, "th percentile")

  # return the results
  return(tibble::tibble(
    observed_value = observed_value,
    het_type = het_type,
    es_type = es_type,
    percentile_range = percentile_range
  ))
}
x <- het_interpret(observed_value = I2_main_model_vector[1],
                   het_type = "I2",
                   es_type = "lnRR_sign",
                   data = intercept_lnRR)
# Sun Aug 31 12:04:36 2025 ------------------------------
# Error in het_interpret(observed_value = I2_main_model_vector[1], het_type = "I2", :
# Invalid effect size type. Check the levels of 'es.type'.
```

@@

```{r sigma2_v intercept_lnRR}
# Then also:
# # Typical sampling variance (which captures the statistical noise of the data, 
# # but is rarely reported in the current meta-analytic practice):
sigma2_v(intercept_lnRR)

@@

# # # Total unstandardized raw variance (i.e. total heterogeneity)
round(sum(intercept_lnRR$sigma2), 3)
```

@@

```{r intercept-only-model-SMDH}
VCV_SMDH <- vcalc(
  vi = SMDH_variance,
  cluster = paper_ID,
  obs = Observation_ID,
  data = dataset_SMDH,
  # assuming that the effect sizes 
  # within the same study are correlated with rho = 0.5
  rho = 0.5 
)

intercept_SMDH <- rma.mv(
  # specify SMDH as the effect size measure;
  yi = SMDH_sign,
  # specify SMDH's sampling variance-covariance matrix;
  V = VCV_SMDH, 
  mods = ~ 1,
  random = list(~ 1 | paper_ID,
                ~ 1 | Observation_ID,
                ~ 1 | group_ID_coded,
                ~ 1 | population_ID,
                ~ 1 | bird_species),
  method = "REML",
  control = list(optimizer="optim"),
  test = "t",
  data = dataset_SMDH)
```

```{r saving SMDH data, eval = FALSE}
## NOT RUN
## -----------------------------------------------------------

## saving the model
saveRDS(intercept_SMDH, file = here::here("model/intercept_SMDH.rds"))

## saving the dataset for this model
write.csv(dataset_SMDH,here::here("data/04_data_analysis/intercept_SMDH.csv"))
```

### SMDH-heterogeneity

@@

# # # Total unstandardized raw variance (i.e. total heterogeneity)
round(sum(intercept_SMDH$sigma2), 3)
```

@@

```{r Bivariate Meta-analytical model}
dataset_bivariate <- data.frame(
  paper_ID = c(dataset_lnRR$paper_ID, dataset_SMDH$paper_ID),
  Observation_ID = c(dataset_lnRR$Observation_ID, dataset_SMDH$Observation_ID),
  group_ID_coded = c(dataset_lnRR$group_ID_coded, dataset_SMDH$group_ID_coded),
  population_ID = c(dataset_lnRR$population_ID, dataset_SMDH$population_ID),
  bird_species = c(dataset_lnRR$bird_species, dataset_SMDH$bird_species),
  ES = c(dataset_lnRR$lnRR_sign, dataset_SMDH$SMDH_sign),
  V_ES = c(dataset_lnRR$lnRR_variance, dataset_SMDH$SMDH_variance),
  ES_measure = c(rep("lnRR", nrow(dataset_lnRR)), rep("SMDH", nrow(dataset_SMDH))))

VCV_bivariate <- vcalc(
  # sampling variances of lnRR and SMD that are correlated within the same study;
  vi = V_ES, 
  # study identity;
  cluster = paper_ID,
  # different types of effect size measures underlying the observed effect sizes;
  type = ES_measure, 
  # the long format data frame;
  data = dataset_bivariate, 
  # assuming that the effect sizes within the same study are correlated with rho = 0.5.
  rho = 0.2, 
  # specify whether impose the non positive definite VCV matrix 
  # (not invertible) to the nearest positive semi-definite matrix
  nearpd = TRUE)

Bivariate_model <- rma.mv(
  # specify the effect size estimate (the variable ES in our case);
  yi = ES, 
  # specify the imputed variance-covariance matrix;
  V = VCV_bivariate, 
  # specify the variable "ES_measure" indicating the types of effect size measures;
  mods = ~ ES_measure - 1, 
  # add correlated random effects corresponding to the lnRR and SMD parameters in the same study;
  random = list(~ ES_measure | paper_ID, 
                ~ ES_measure | Observation_ID,
                ~ 1 | group_ID_coded,
                ~ 1 | population_ID,
                ~ 1 | bird_species
  ), 
  # impose an unstructured variance-covariance of the study-specific random effects;
  struct = "UN",
  # t distribution is specified to test the overall effect against the null hypothesis and construct confidence intervals;
  test = "t", 
  # restricted likelihood maximum is assigned as the estimator for variance components as suggested;
  method = "REML", 
  # the long format dataset
  data = dataset_bivariate, 
  sparse = T)

@@

```{r, eval = FALSE}
## NOT RUN
## -----------------------------------------------------------

# Saving the model
saveRDS(Bivariate_model, 
        file = here::here("model/sensitivity_analysis/bivariate_model.rds"))

# saving the dataset for this model
write.csv(dataset_bivariate,
          here::here("data/05_data_sensitivity_analysis/dataset_bivariate.csv"))
```

```{r WARNING = F}
# fit regression model
ab.l <- matreg(
  y = 2, 
  x = 1,
  R = Bivariate_model$G, 
  cov = TRUE, 
  means = coef(Bivariate_model), 
  n = Bivariate_model$g.levels.comb.k)

xy <- ellipse(Bivariate_model$G, 
              centre = coef(Bivariate_model),
              # get 95% coverage region
              level = 0.95)

## convert it into data frame 
## so that we can use ggplot to make a figure
ellipse_df <- data.frame(x = xy[, 1],
                         y = xy[, 2])

@@

There is a very high negative correlation between the two effect sizes. This probably is an artefact of the data and not a true correlation. If you look at the model summary, the $tau^2$ for lnRR, it is extremely small 0.0008, i.e., there is very little variance. This could also be stemming from the fact that the log values in the dataset could be causing this. For SMDH on the other hand, the $tau^2$ is 0.1681 that is much larger. This difference in the variance estimate itself could be causing this artefact of the model. I will only go ahead to present this model as a sensitivity analysis because I am not completely convinced by it's robustness.

## Sensitivity Analysis

@@

# Preparing new variance-covariance matrix
VCV_lnRR_published <- vcalc(
  vi = lnRR_variance,
  cluster = paper_ID,
  obs = Observation_ID,
  data = dataset_lnRR_published,
  # assuming that the effect sizes within the same study
  #  are correlated with rho = 0.5
  rho = 0.5 
)

## Running intercept only model
SA_lnRR_published <- rma.mv(
  # specify lnRR as the effect size measure;
  yi = lnRR_sign, 
  # specify lnRR's sampling variance-covariance matrix;
  V = VCV_lnRR_published, 
  mods = ~ 1,
  random = list(~ 1 | paper_ID,
                ~ 1 | Observation_ID,
                ~ 1 | group_ID_coded,
                ~ 1 | population_ID,
                ~ 1 | bird_species),
  method = "REML",
  control = list(optimizer="optim"),
  test = "t",
  data = dataset_lnRR_published)
```

```{r, eval = F}
## NOT RUN
## -----------------------------------------------------------

# Saving the model
saveRDS(SA_lnRR_published, 
        file = here::here("model/sensitivity_analysis/SA_lnRR_published.rds"))

# saving the dataset for this model
write.csv(dataset_lnRR_published,
          here::here("data/05_data_sensitivity_analysis/dataset_lnRR_published.csv"))
```

```{r}
# What is the effect now?
summary(SA_lnRR_published)
predict(SA_lnRR_published)

# Preparing new variance-covariance matrix
VCV_SMDH_published <- vcalc(
  vi = SMDH_variance,
  cluster = paper_ID,
  obs = Observation_ID,
  data = dataset_SMDH_published,
  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
)

## Running intercept only model
SA_SMDH_published<- rma.mv(
  yi = SMDH_sign, # specify lnRR as the effect size measure;
  V = VCV_SMDH_published, # specify lnRR's sampling variance-covariance matrix;
  mods = ~ 1,
  random = list(~ 1 | paper_ID,
                ~ 1 | Observation_ID,
                ~ 1 | group_ID_coded,
                ~ 1 | population_ID,
                ~ 1 | bird_species),
  method = "REML",
  control = list(optimizer="optim"),
  test = "t",
  data = dataset_SMDH_published)
```

```{r, eval =FALSE}
## NOT RUN
## -----------------------------------------------------------

## Saving the model
saveRDS(SA_SMDH_published, 
        file = here::here("model/sensitivity_analysis/SA_SMDH_published.rds"))

## saving the dataset for this model
write.csv(dataset_SMDH_published,
          here::here("data/05_data_sensitivity_analysis/dataset_SMDH_published.csv"))
```

```{r}
# What is the effect now?
summary(SA_SMDH_published)
predict(SA_SMDH_published)
```
